---
title: "main"
author: "Harrison Burke"
date: "12/3/2019"
output:
  pdf_document: default
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(shiny)
library(plotly)
library(shinyWidgets)
library(fs)
library(shinythemes)
library(moderndive)
library(gt)
library(tidyverse)
library(countrycode)

#reading in csv
data <- read.csv("final-project/data/athlete_events.csv")

#preparing data for cleaning
row_data <- data %>%
  filter(Sport == "Rowing") %>%
  mutate(Medal = as.character(Medal))

#turning all NA values to 0
row_data[is.na(row_data)] <- 0


download.file(
  "https://raw.githubusercontent.com/datasets/population/master/data/population.csv",
  destfile = "final-project/data/pop.csv"
)

world_pop <- read_csv("final-project/data/pop.csv")

# Changing variable name and format so I can merge with the rowing dataset
clean_pop <- world_pop %>%
  rename(NOC = "Country Code", total_pop = Value) %>%
  mutate(NOC = countrycode(NOC, origin = "iso3c", destination = "ioc"))


# Joining rowing data with population data
row_data_pop <-
  inner_join(clean_pop, row_data, by = c("NOC", "Year"))


# Reading in gsp data
gdp <- read_csv("final-project/data/gdp.csv")


# Reformatting and renaming data so I can merge with rowing + population dataset. I had to write in the name of every column in that nasty format,
# which I then cleaned up using str_extract to get rid of the square brackets
clean_gdp <- gdp %>%
  select(-"Series Name",-"Series Code") %>%
  pivot_longer(
    cols = c(
      "1960 [YR1960]",
      "1961 [YR1961]",
      "1962 [YR1962]",
      "1963 [YR1963]",
      "1964 [YR1964]",
      "1965 [YR1965]",
      "1966 [YR1966]",
      "1967 [YR1967]",
      "1968 [YR1968]",
      "1969 [YR1969]",
      "1970 [YR1970]",
      "1971 [YR1971]",
      "1972 [YR1972]",
      "1973 [YR1973]",
      "1974 [YR1974]",
      "1975 [YR1975]",
      "1976 [YR1976]",
      "1977 [YR1977]",
      "1978 [YR1978]",
      "1979 [YR1979]",
      "1980 [YR1980]",
      "1981 [YR1981]",
      "1982 [YR1982]",
      "1983 [YR1983]",
      "1984 [YR1984]",
      "1985 [YR1985]",
      "1986 [YR1986]",
      "1987 [YR1987]",
      "1988 [YR1988]",
      "1989 [YR1989]",
      "1990 [YR1990]",
      "1991 [YR1991]",
      "1992 [YR1992]",
      "1993 [YR1993]",
      "1994 [YR1994]",
      "1995 [YR1995]",
      "1996 [YR1996]",
      "1997 [YR1997]",
      "1998 [YR1998]",
      "1999 [YR1999]",
      "2000 [YR2000]",
      "2001 [YR2001]",
      "2002 [YR2002]",
      "2003 [YR2003]",
      "2004 [YR2004]",
      "2005 [YR2005]",
      "2006 [YR2006]",
      "2007 [YR2007]",
      "2008 [YR2008]",
      "2009 [YR2009]",
      "2010 [YR2010]",
      "2011 [YR2011]",
      "2012 [YR2012]",
      "2013 [YR2013]",
      "2014 [YR2014]",
      "2015 [YR2015]",
      "2016 [YR2016]",
      "2017 [YR2017]",
      "2018 [YR2018]",
      "2019 [YR2019]"
    ),
    names_to = "Year",
    values_to = "gdp_percap"
  ) %>%
  mutate(Year = str_extract(Year, "^\\d{4}")) %>%
  mutate(Year = as.integer(Year)) %>%
  rename(NOC = "Country Code") %>%
  mutate(NOC = countrycode(NOC, origin = "iso3c", destination = "ioc"))


# Joining rowing dataset that includes population data with the gdp dataset
big_data <-
  inner_join(clean_gdp, row_data_pop, by = c("NOC", "Year"))


model_data_gdp <- big_data %>%
  mutate(gdp_percap = as.double(gdp_percap)) %>%
  group_by(Year, Event, NOC) %>%
  distinct(Event, .keep_all = TRUE) %>%
  add_count(Medal) %>%
  pivot_wider(
    names_from = "Medal",
    values_from = "n",
    values_fill = list(n = 0)
  ) %>%
  group_by(Year, total_pop, gdp_percap) %>%
  mutate(ycount = Gold + Bronze + Silver) %>%
  group_by(Year, total_pop, gdp_percap, NOC) %>%
  summarize(
    Gold = sum(Gold),
    Bronze = sum(Bronze),
    Silver = sum(Silver),
    Total = sum(ycount)
  ) %>%
  pivot_longer(
    cols = c(Gold, Bronze, Silver, Total),
    names_to = "Medal",
    values_to = "nmedals"
  ) %>%
  filter(Medal == "Total") %>%
  mutate(medals_per_mil = (nmedals / total_pop) * 1000000)

```

## The research question(s)

What is the relationship between rowing success and economic success?

## Background/significance of the research

My goal in this project was to take a deeper dive into factors that influence rowing success at the Olympics. The stereotype of the sport is that it's elitist and only for the wealthy, so I made a point to look at the relationship between wealth metrics and rowing results. Furthermore, there is a surprising lack of interesting rowing data on the web, so even exploring within the dataset will uncover some interesting stories.

## The methods used to obtain and analyze the data

I completed the project entirely in R. I downloaded datasets from Kaggle and the World Banks, and mainly used the tidyverse library to clean, transform, and merge the data. Once aggregated, I used many different libraries to help me manipulate the data to get the information I wanted. A list of libraries that I used will be attached here:

shiny
plotly
shinyWidgets
fs
shinythemes
moderndive
gt
tidyverse
countrycode

I performed standard linear regression with GDP Per Capita as the explanatory variable, and the medal count at the olympics for the dependent variable. I visualized this relationship with a scatter plot and a regression line, and I printed the output of the model to a table.




## A discussion of the research, the limitations of the current research, reasonableness of any assumptions made, possibilities of future work/studies that should be conducted, etc.

The research was fascinating, as I had not explored data of this type before. I was somewhat surprised in the variability of the model between countries, but considering how many factors there are to explore, I also knew that there was no way it was going to be a perfect fit. I was focused on the data that I had, but there are many more factors that I would like to explore. For instance, if the percentage of a country that is water has any effect, the number of rowing clubs, height, weight, skin color. The more factors that can be explored, the better the understanding will be from a meta-perspective.




```{r, echo=FALSE}


countries_with_medals <- row_data %>%
    mutate(medal_id = case_when(
        Medal == "0" ~ 0,
        Medal == "Bronze" ~ 1,
        Medal == "Silver" ~ 2,
        Medal == "Gold" ~ 3
    )) %>%
    group_by(NOC) %>%
    summarize(medal_check = sum(medal_id)) %>%
    filter(medal_check > 0) %>%
    mutate(NOC = as.character(NOC))
           
  

row_data %>%
  filter(NOC == "USA") %>%
  group_by(Year, Event, NOC) %>%
  distinct(Event, .keep_all = TRUE) %>%
  add_count(Medal) %>%
  select(Year, NOC, Event, Medal, n) %>%
  pivot_wider(
    names_from = "Medal",
    values_from = "n",
    values_fill = list(n = 0)
  ) %>%
  group_by(Year) %>%
  mutate(ycount = Gold + Bronze + Silver) %>%
  group_by(Year) %>%
  summarize(
    Gold = sum(Gold),
    Bronze = sum(Bronze),
    Silver = sum(Silver),
    Total = sum(ycount)
  ) %>%
  pivot_longer(
    cols = c(Gold, Bronze, Silver, Total),
    names_to = "Medal",
    values_to = "nmedals"
  ) %>%
  ggplot(aes(x = Year, y = nmedals, color = Medal)) + geom_line(size = 1) +
  scale_color_manual(values = c("#cd7f32", "#d4af37", "#aaa9ad", "#000000")) +
  expand_limits(y = c(0, 15)) +
  theme_minimal()

```

## The results of the analysis (tables, charts, graphs, significance, confidence intervals, descriptive text)

Here, I am only looking at the data for the United States. As is clear in the graph, there is not a strong relationship between GDP per capita and the number of medals won at the olympics.
```{r, echo=FALSE}
# Medals and gdp per capita

big_data %>%
            mutate(gdp_percap = as.double(gdp_percap)) %>%
            group_by(Year, Event, NOC) %>%
            distinct(Event, .keep_all = TRUE) %>%
            add_count(Medal) %>%
            # select(Year, NOC, Event, Medal, n) %>%
            pivot_wider(
                names_from = "Medal",
                values_from = "n",
                values_fill = list(n = 0)
            ) %>%
            filter(NOC == "USA") %>%
            group_by(Year, total_pop, gdp_percap) %>%
            mutate(ycount = Gold + Bronze + Silver) %>%
            group_by(Year, total_pop, gdp_percap) %>%
            summarize(
                Gold = sum(Gold),
                Bronze = sum(Bronze),
                Silver = sum(Silver),
                Total = sum(ycount)
            ) %>%
            pivot_longer(
                cols = c(Gold, Bronze, Silver, Total),
                names_to = "Medal",
                values_to = "nmedals"
            ) %>%
            filter(Medal == "Total") %>%
            group_by(Year, gdp_percap) %>%
            mutate(medals_per_mil = (nmedals / total_pop) * 1000000) %>%
            ggplot(aes(x = gdp_percap, y = nmedals)) + geom_point() + geom_smooth(method = "lm", se = F) + ggtitle("The Relationship Between GDP and Number of Medals") +
            labs(y = "Number of Medals Won") +
            theme_minimal()
```

The table more clearly describes the model. The model predicts that for the United States, when GDP is 0, the number of medals that will be won will be around 3. The table is helped by the graph. The project allows users to manipulate the country that they are looking at, so the data ostensibly varies from country to country. Some countries have a strong positive correlation, others have a strong negative correlation, and many, like the US, have little to no correlation. I expected more consistency across countries, but there are obviously many more factors at play than what is accounted for by GDP Per Capita.

```{r, echo=FALSE}
## GDP Per Capita Model


gdp_percap_model_data <- big_data %>%
  mutate(gdp_percap = as.double(gdp_percap)) %>%
  group_by(Year, Event, NOC) %>%
  distinct(Event, .keep_all = TRUE) %>%
  add_count(Medal) %>%
  pivot_wider(
    names_from = "Medal",
    values_from = "n",
    values_fill = list(n = 0)
  ) %>%
  group_by(Year, total_pop, gdp_percap) %>%
  mutate(ycount = Gold + Bronze + Silver) %>%
  group_by(Year, total_pop, gdp_percap, NOC) %>%
  summarize(
    Gold = sum(Gold),
    Bronze = sum(Bronze),
    Silver = sum(Silver),
    Total = sum(ycount)
  ) %>%
  pivot_longer(
    cols = c(Gold, Bronze, Silver, Total),
    names_to = "Medal",
    values_to = "nmedals"
  ) %>%
  filter(Medal == "Total") %>%
  mutate(medals_per_mil = (nmedals/total_pop)*1000000) %>%
  select(NOC, Year, gdp_percap, nmedals) %>%
  filter(NOC == "USA")

        
model_gdp <-
  get_regression_table(lm(nmedals ~ gdp_percap, data = gdp_percap_model_data))

# Creating table
model_gdp %>%
  select(term, estimate, lower_ci, upper_ci) %>%
  mutate(term = c("Intercept", "Height")) %>%
  gt() %>%
  cols_label(
    term = "",
    estimate = "Coefficient",
    lower_ci = "5th percentile",
    upper_ci = "95th percentile"
  ) %>%
  fmt_number(columns = 2:4,
             decimals = 3) %>%
  tab_header(title = "Country GDP Per Capita and Number of Medals Won",
                       subtitle = "The relationship between GDP per capita and medal count varies by country")

```



## Most Medals per Million people at the 2016 Olympics

```{r, echo=FALSE}
# Medals Per Cap by Country table

big_data %>%
  mutate(NOC = countrycode(NOC, origin = "ioc", destination = "country.name")) %>%
  mutate(gdp_percap = as.double(gdp_percap)) %>%
  group_by(Year, Event, NOC, total_pop) %>%
  distinct(Event, .keep_all = TRUE) %>%
  add_count(Medal) %>%
  pivot_wider(
    names_from = "Medal",
    values_from = "n",
    values_fill = list(n = 0)
  ) %>%
  filter(Year == "2016") %>%
  group_by(NOC, total_pop) %>%
  summarize(total = sum(Gold, Bronze, Silver)) %>%
  ungroup() %>%
  mutate(medals_per_mil = (total/total_pop)*1000000) %>%
  select(NOC, medals_per_mil) %>%
  arrange(desc(medals_per_mil)) %>%
  gt() %>%
  cols_label(
    NOC = "Country",
    medals_per_mil = "# of Medals per Million People"
  )


```

## References

https://raw.githubusercontent.com/datasets/population/master/data/population.csv

https://www.kaggle.com/heesoo37/120-years-of-olympic-history-athletes-and-results






